### 2024-03-05 tex/syntax.tex
[ correction ] first pass over chapter 6

The \AgdaDatatype{Premise} data type describes premises of rules,
using the bunched connectives from \cref{fig:bunched}. A single
premise is introduced by the
\AgdaInductiveConstructor{$\langle$\_`$\vdash$\_$\rangle$}
constructor.  This allows binding of additional variables
\AgdaBound{$\Delta$} (with specified types and usage annotations) and
the specification of a conclusion type \AgdaBound{A} for this premise.
The remaining constructors are descriptions for the
bunched connectives. %, and will be interpreted directly as such, below.

@@

Finally, a \AgdaRecord{System} consists of a set of rule labels (i.e.,
constructor names), and for each label a description of the
corresponding rule. We use $\rhd$ as infix notation for systems to
associate the label set with the rules.

@@

As an example, we transcribe a fragment of $\name$ (as defined in
\cref{fig:lr} and \cref{fig:lr-bunched}) into a description.
We give the set of types of
this system as a data type \AgdaDatatype{Ty} (together with a base
type \AgdaInductiveConstructor{$\iota$}). We assume that there is a
posemiring \AgdaInductiveConstructor{Ann} in scope for the
annotations.
There is one label for each instantiation of a logical
rule, but the labels contain no further information about subterms or
restrictions on the context. This will be provided when we associate
labels with \AgdaRecord{Rule}s in a \AgdaRecord{System}.

@@

Compared to \cref{fig:lr-bunched}, modulo the Agda notation, we can see
that the fundamental structure has been preserved: The rules match
one-to-one, and the bunched premises are the same. A major difference
is that we do not include a counterpart to the
\AgdaInductiveConstructor{var} rule in a
\AgdaRecord{System}. Variables are common to all the systems
representable in our framework.

@@

The next thing we want to do is to build terms in the described type system.
The following definitions are useful for talking about types indexed over
contexts, judgement forms, and judgement forms admitting newly bound variables,
respectively.

@@

To specify the meaning of descriptions, we assume some
\AgdaBound{X}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}\AgdaFunction{ExtOpenFam}%
\AgdaSpace{}\AgdaSymbol{\_},
% \ExecuteMetaData[\Interpretationtex]{X},
over which we form one layer of syntax, using the function
\AgdaFunction{$\llbracket$\_$\rrbracket$p} that interprets
\AgdaDatatype{Premises} defined below.  The first argument to
\AgdaBound{X} is the new variables bound by this layer of syntax, as
exemplified in the first clause of
\AgdaFunction{$\llbracket$\_$\rrbracket$p}.  The second argument is
the context containing the variables being carried over from the
previous layer.  Notice that this is not, in general, the same as the
context from the previous layer, because the usage annotations may
have been changed by connectives like
\AgdaInductiveConstructor{\_`$*$\_} and
\AgdaInductiveConstructor{\_`$\cdot$\_}.  The third argument is the
type of subterm required.

@@

% A layer of syntax supports the following functorial action.

% \ExecuteMetaData[\Maptex]{map-s-type}

@@

\subsection{Duplicability and L/nL}\label{sec:dup-lnl}
In \cref{sec:rec}, I introduced a bunched connective $\Box^{0{+}}$, used to
ensure that such a premise is derived in a context that allows it to be
duplicated and discarded.
Syntax for this connective can be added to \AgdaDatatype{Premises} and worked
through the various parts of the framework, though I do not show these parts in
the text of this thesis for the sake of simplicity and brevity.
Additionally, variations such as $\Box^0$ (requiring only that the context is
${} \leq \gr0$) and $\Box^{0{+}{*}}$ (requiring, in addition to
$\Box^{0{+}}$, closure under multiplication by any scalar) can be added to the
framework independently of each other and of $\Box^{0{+}}$.

@@

The rules are presented in Agda code below, and also using bunched inference
rule notation in \cref{fig:LnL-bunched}.

>>>>

The \AgdaDatatype{Premises} data type describes premises of rules,
using the bunched connectives from \cref{fig:bunched}. A single
premise is introduced by the
\AgdaInductiveConstructor{$\langle$\_`$\vdash$\_$\rangle$}
constructor.  This allows binding of additional variables
\AgdaBound{$\Delta$} (with specified types and usage annotations) and
the specification of a conclusion type \AgdaBound{A} for this premise.
The remaining constructors are descriptions for the
bunched connectives. %, and will be interpreted directly as such, below.

@@

Finally, a \AgdaRecord{System} consists of a set of rule labels (i.e.,
constructor names), and for each label a description of the
corresponding rule. I use $\rhd$ as infix notation for systems to
associate the label set with the rules.

@@

As an example, I transcribe a fragment of $\name$ (as defined in
\cref{fig:lr} and \cref{fig:lr-bunched}) as a description.
I give the set of types of
this system as a data type \AgdaDatatype{Ty} (together with a base
type \AgdaInductiveConstructor{$\iota$}).
I assume that there is a
posemiring \AgdaInductiveConstructor{Ann} in scope for the
annotations.
There is one label for each instantiation of a logical
rule, but the labels contain no further information about subterms or
restrictions on the context.
This will be provided by the \AgdaRecord{Rule} associated to each label in a
\AgdaRecord{System}.

@@

Compared to \cref{fig:lr-bunched}, modulo the Agda notation, we can see
that the fundamental structure has been preserved: The rules match
one-to-one, and the bunched premises are the same. A major difference
is that I do not include a counterpart to the
\AgdaInductiveConstructor{var} rule in a
\AgdaRecord{System}. Variables are common to all the systems
representable in this framework.

@@

The next thing we want to do is to build terms in the described type system.
The following three definitions are useful for talking about types indexed over
contexts.
An \emph{open type} (\AgdaFunction{OpenType}) is a type
indexed over a context.
These are the things that bunched connectives connect, and thus serve as the
interpretation of \AgdaDatatype{Premises}.

\ExecuteMetaData[\Syntaxtex]{OpenType}

An \emph{open family} (\AgdaFunction{OpenFam}) is
indexed on a context and additionally on an object language type.
Judgement forms, such as the families of variables, terms, and semantics
thereof, are examples of open families.

@@

An \emph{extensible open family} (\AgdaFunction{ExtOpenFam})
is like an open family, but is indexed over an additional context.
This additional context is intended to represent newly bound variables, which
may just be added to the existing context, or the extensible open family may
treat them differently to the existing variables.

\ExecuteMetaData[\Syntaxtex]{ExtOpenFam}

To specify the meaning of descriptions, let us assume some
extensible open family \AgdaBound{X},
over which we form one layer of syntax, using the function
\AgdaFunction{$\llbracket$\_$\rrbracket$p} that interprets
\AgdaDatatype{Premises} defined below.  The first argument to
\AgdaBound{X} is the new variables bound by this layer of syntax, as
exemplified in the first clause of
\AgdaFunction{$\llbracket$\_$\rrbracket$p}.  The second argument is
the context containing the variables being carried over from the
previous layer.  Notice that this is not, in general, the same as the
context from the previous layer, because the usage annotations may
have been changed by connectives like
\AgdaInductiveConstructor{\_`$*$\_} and
\AgdaInductiveConstructor{\_`$\cdot$\_}.  The third argument is the
type of subterm required.

@@

@@

\subsection{Duplicability and L/nL}\label{sec:dup-lnl}
In \cref{sec:rec}, I introduced a bunched connective $\Box^{0{+}}$, used to
ensure that such a premise is derived in a context that allows it to be
duplicated and discarded.
Syntax for this connective can be added to \AgdaDatatype{Premises} and worked
through the various parts of the framework, though I do not show these parts in
the text of this thesis for the sake of simplicity and brevity.
Additionally, variations such as $\Box^0$ (requiring only that the context is
${} \leq \gr0$) and $\Box^{0{+}{*}}$ (requiring, in addition to
$\Box^{0{+}}$, closure under multiplication by any scalar) can be added to the
framework independently of each other and of $\Box^{0{+}}$.
I define $\Box^{0{+}{*}}$ as follows.

\[
  \plr{\Box^{0{+}{*}}T}\,\grP\gamma \coloneqq
  \plr{\grP \leq \grQ} \times
  \plr{\grQ \leq \gr0} \times \plr{\grQ \leq \grQ + \grQ} \times
  \plr{\forall\gr r.~\grQ \leq \gr r\grQ} \times T\,\grQ\gamma
\]

@@

The rules are presented in Agda code below, and also using bunched inference
rule notation in \cref{fig:LnL-bunched}.
Recall that \AgdaDatatype{Hand} is a type containing two inhabitants ---
\AgdaInductiveConstructor{ll} and \AgdaInductiveConstructor{rr}.
I use the notation \AgdaFunction{[} \AgdaBound{X} \AgdaFunction{>}
\AgdaBound{i} \AgdaFunction{<} \AgdaBound{Y} \AgdaFunction{]}, for
\AgdaBound{i} \AgdaSymbol: \AgdaDatatype{Hand}, to mean \AgdaBound{X} when
\AgdaBound{i} is \AgdaInductiveConstructor{ll} and \AgdaBound{Y} when
\AgdaBound{i} is \AgdaInductiveConstructor{rr}.
